Static contract checking for Haskell

### 2011-08-17 paper.tex
introduction

We start by stating that $Nil$ and $Cons$ can never be equal:
$$\forall a,b.~Cons(a,b) \neq Nil$$

@@

$$\cf{Nil}$$
$$\forall a,b.~\cf{a} \land \cf{b} \iff \cf{Cons(a,b)}$$

We also say some stuff about unreachibility but I can't think of a
good way to explain it right now.
$$\forall y,ys.~Cons(y,ys) \neq \unr $$
$$Nil \neq \unr $$

Finally, we define projections for $Cons$. It is not strictly
necessary, but it will be handy:
$$\forall xs,y,ys.~sel_{1,Cons}(Cons(y,ys)) = xs \implies xs = y $$
$$\forall xs,y,ys.~sel_{2,Cons}(Cons(y,ys)) = xs \implies xs = ys $$

Now we translate the $null$ function. Note that the symbols $true$ and
$false$ are the representation of the data constructors $True$ and
$False$ in Haskell, not the boolean values $\top$ and $\bot$ in our
logic.
$$\forall xs.~ xs=Nil \implies notnull(xs) = false$$
$$\forall xs,y,ys.~ x=Cons(y,ys) \implies notnull(xs) = true$$

We also need to specify the translation of calls to $notnull$ with
$\bad$ values (to encode the fact that $notnull(\bad) = \bad$.
$$\forall xs.~xs=\bad \to notnull(xs) = \bad$$

Finally, we say that one call $notnull$ with a an argument which is
not $Nil$ or $Cons$ or $\bad$ then the result is $\unr$.
$$\forall xs.~xs \neq \bad \land xs \neq Nil$$
$$\land xs \neq Cons(sel_{1,Cons}(xs),sel_{2,Cons}(xs)) \to $$
$$ notnull(X) = \unr$$

The translation of $head$ follows the same pattern:
$$\forall xs.~ x=Nil \implies head(xs) = \bad$$
$$\forall xs,y,ys.~ x=Cons(y,ys) \implies head(xs) = y$$

$$\forall xs.~xs=\bad \to head(xs) = \bad$$

$$\forall xs.~xs \neq \bad \land xs \neq Nil$$
$$ \land xs \neq Cons(sel_{1,Cons}(xs),sel_{2,Cons}(xs)) \to $$
$$ head(X) = \unr$$

We now have translated the source code. Let us call all those formulae
the theory $T$. We translate separatly the contract:
$$ \phi := \forall xs.~\cf{xs} \land not(null(xs)) = true \to \cf{head(xs)} $$

@@

Intuitivly, $T$ is consistent (ie $T \not \vdash \bot$), because each
formula serves a specific purpose. Now, assume that $xs$ satisfies
$\cf{xs}$ and $notnull(xs) = true$.  We can derive that $xs \neq \bad$
because we have $\cf{xs}$ and $\not \cf{\bad}$. The constraint
$notnull(xs) = true$ doesn't directly imply that $xs = Cons(y,ys)$ for
some $y$ and $ys$. But the translation of $notnull$ is total, ie the
function is totally defined. This implies that there exist $y$ and
$ys$ such that $xs = Cons(y,ys)$. Recalling $\cf{xs}$, we can now
derive $\cf{y}$ and $\cf{ys}$. But $head(xs) = y$ in our setting, and
$y$ is crash-free. QED.

>>>>

We start by stating that $Nil$ and $Cons$ can never be equal:
\begin{equation*}
\forall a,b.~Cons(a,b) \neq Nil
\end{equation*}

@@

\begin{equation*}
\cf{Nil}
\end{equation*}
\begin{equation}
  \label{cfcons}
\forall a,b.~\cf{a} \land \cf{b} \iff \cf{Cons(a,b)}
\end{equation}

We also say some stuff about unreachibility but I can't think of a
good way to explain it right now.
\begin{equation*}
\forall y,ys.~Cons(y,ys) \neq \unr 
\end{equation*}
\begin{equation*}
Nil \neq \unr 
\end{equation*}

Finally, we define projections for $Cons$. It is not strictly
necessary, but it will be handy:
\begin{equation**}
\forall xs,y,ys.~sel_{1,Cons}(Cons(y,ys)) = xs \implies xs = y 
\end{equation**}
\begin{equation*}
\forall xs,y,ys.~sel_{2,Cons}(Cons(y,ys)) = xs \implies xs = ys 
\end{equation*}

Now we translate the $null$ function. Note that the symbols $true$ and
$false$ are the representation of the data constructors $True$ and
$False$ in Haskell, not the boolean values $\top$ and $\bot$ in our
logic.
\begin{equation*}
\forall xs.~ xs=Nil \implies notnull(xs) = false
\end{equation*}

\begin{equation}
  \label{defnotnull}
\forall xs,y,ys.~ x=Cons(y,ys) \implies notnull(xs) = true
\end{equation}

We also need to specify the translation of calls to $notnull$ with
$\bad$ values (to encode the fact that $notnull(\bad) = \bad$.
\begin{equation*}
\forall xs.~xs=\bad \to notnull(xs) = \bad
\end{equation*}

Finally, we say that one call $notnull$ with a an argument which is
not $Nil$ or $Cons$ or $\bad$ then the result is $\unr$.
\begin{align}
  \label{unrnotnull}
  \forall xs.~xs \neq \bad \land xs \neq Nil  \notag \\
  \land xs \neq Cons(sel_{1,Cons}(xs),sel_{2,Cons}(xs)) \to  \notag \\
  notnull(X) = \unr 
\end{align}

The translation of $head$ follows the same pattern:
\begin{align}
  \forall xs.~ x=Nil \implies head(xs) = \bad \notag\\
  \forall xs,y,ys.~ x=Cons(y,ys) \implies head(xs) = y     \label{defhead} \\
    \forall xs.~xs=\bad \to head(xs) = \bad \notag
\end{align}

\begin{align*}
 \forall xs.~xs \neq \bad \land xs \neq Nil \notag \\
 \land xs \neq Cons(sel_{1,Cons}(xs),sel_{2,Cons}(xs)) \to  \notag \\
 head(X) = \unr 
\end{algin*}

We now have translated the source code. Let us call all those formulae
the theory $T$. We translate separatly the contract:
\begin{equation*}
  \phi := \forall xs.~\cf{xs} \land notnull(xs) = true \to \cf{head(xs)} 
\end{equation*}

@@

Intuitively, $T$ is consistent (ie $T \not \vdash \bot$), because each
formula serves a specific purpose. Now, assume that $xs$ satisfies
$\cf{xs}$ and $notnull(xs) = true$.  We can derive that $xs \neq \bad$
because we have $\cf{xs}$ and $\lnot \cf{\bad}$. The constraint
$notnull(xs) = true$ doesn't directly imply that $xs = Cons(y,ys)$ for
some $y$ and $ys$. But $notnull$ is totally defined, because of
\eqref{unrnotnull}. This implies (by \eqref{defnotnull}) that there exist
$y$ and $ys$ such that $xs = Cons(y,ys)$. Recalling $\cf{xs}$, we can
now derive $\cf{y}$ and $\cf{ys}$ (by \eqref{cfcons}). But $head(xs) =
y$ because of \eqref{defhead}, and $y$ is crash-free, so we can finally
derive $\cf{head(xs)}$. QED.
